manuscript

### 2023-12-20 manuscript/ms.Rmd
Now with short, focused discussion

The average allele count per locus (which is usually corrected to allelic richness [@kalinowskiCountingAllelesRarefaction2004]) has historically been one of the most prevalent allele frequency independent measures of genetic diversity. While this is a particularly useful measure in microsatellite and other datasets utilizing heavily poly-allelic markers, it is less so in single-nucleotide polymorphism (SNP) datasets. SNP datasets are usually primarily biallelic, and thus the count of the number of alleles per locus in such datasets is less variable across populations and therefore less interesting. Counts of the number of segregating sites per population are a useful alternative in SNP datasets given the large number of possible segregating sites that these datasets often query [@?CITE].

@@

<!-- Outline: -->

<!-- * What is genetic diveristy? -->
<!--   - Utility -->
<!--   - Estimators -->
<!--     - heterozygosity -->
<!--     - pi -->
<!--     - private alleles -->
<!--     - allele counts per locus -->

<!-- * Problems with ac and private alleles per locus -->
<!--   - All subject to bias due to differences in effective sample size -->
<!--     - Missing data, unequal sample sizes -->

<!-- * Rarefaction -->
<!--   - Sampling vs mathematical solutions -->
<!--   - solutions for private alleles and allele counts (richness) -->
<!--   - description -->
<!--   - how this solves problems -->

<!-- * Nseg -->
<!--   - Newer metric for modern sequencing tech -->
<!--   - Useful alternative, lets you capture different parts of variance -->
<!--   - biases -->
<!--   - rarefaction as solution -->

@@

To remedy these problems, I propose the following estimator of $P({S_{j}})$:
$$P({S_{j}}) = 1 - \sum_{k = 1}^{h}\frac{\binom{n_{kj}}{\gamma}}{\binom{n_{j}}{\gamma}}$$
where the $P({S_{j}})$ is given by the probability of exclusively drawing any $k$ of $h$ possible homozygote genotypes in population $j$ given $\gamma$ independent sampled *genotypes* (not *gene copies*) from the pool of observed genotypes. Here, $n_{kj}$ is the number of observed homozygote genotypes of type $k$ in population $j$ and $n_{j}$ is the total number of observed genotypes of all types. Note that $n_{j}$ and $\gamma$ will be half the value of their equivalents $N_{j}$ and $g$ for diploid species, one third for triploids, and so on.

@@

## Emperical Validation
To validate equations 4 and 5, I simulated genotypic data for two populations with sizes 100 and 1000, each with 100 bi-allelic loci with minor allele frequencies spaced equally between 0.01 and 0.1. I added missing data to each population assigning each locus a missing data rate $R_{mq}$ from a uniform distribution such that $R_{mq} \sim U(0,.3)$, ensuring that overall allele frequencies in each population were maintained. I then used the methods described above to estimate $P({S_{jq}})$ and $E(N_{S})$ and their variances given $\gamma = 30$. For comparison, I also conducted 10,000 random draws of size $\gamma$ from each loci in each population, then calculated $P({S_{jq}})$ and its variance for each locus empirically and $E(N_{S})$ by summing across all 100 loci for each set of draws. I likewise calculated the variance of $N_{S}$ directly across all sets of random draws.

To compare my calculated estimates to the empirical simulations, I used the implementation of the Agresti-Coull [@agrestiApproximateBetterExact1998] method from the R package `binom` [@dorai-rajBinomBinomialConfidence2022] to calculate 95% confidence intervals for each empirical $P({S_{jq}})$. I also used normal approximation to calculate 95% confidence and prediction intervals for $E(N_{S})$ and $N_{S}$ using $\sigma_{E(n_{S})}^{2}$ and the variance directly observed from the simulations, respectively.

@@

# Discussion

>>>>

The average allele count per locus (which is usually corrected to allelic richness [@kalinowskiCountingAllelesRarefaction2004]) has historically been one of the most prevalent allele frequency independent measures of genetic diversity. While this is a particularly useful measure in microsatellite and other datasets utilizing heavily polyallelic markers, it is less so in single-nucleotide polymorphism (SNP) datasets. SNP datasets are usually primarily biallelic, and thus the count of the number of alleles per locus in such datasets is less variable across populations and therefore less interesting. Counts of the number of segregating sites per population are a useful alternative in SNP datasets given the large number of possible segregating sites that these datasets often query [@?CITE].

@@

@@

To remedy these problems, we propose the following estimator of $P({S_{j}})$:
$$P({S_{j}}) = 1 - \sum_{k = 1}^{h}\frac{\binom{n_{kj}}{\gamma}}{\binom{n_{j}}{\gamma}}$$
where the $P({S_{j}})$ is given by the probability of exclusively drawing any $k$ of $h$ possible homozygote genotypes in population $j$ given $\gamma$ independent sampled *genotypes* (not *gene copies*) from the pool of observed genotypes. Here, $n_{kj}$ is the number of observed homozygote genotypes of type $k$ in population $j$ and $n_{j}$ is the total number of observed genotypes of all types. Note that $n_{j}$ and $\gamma$ will be half the value of their equivalents $N_{j}$ and $g$ for diploid species, one third for triploids, and so on.

@@

## Emperical Validation
To validate equations 4 and 5, we simulated genotypic data for two populations with sizes 100 and 1000, each with 100 bi-allelic loci with minor allele frequencies spaced equally between 0.01 and 0.1. We added missing data to each population assigning each locus a missing data rate $R_{mq}$ from a uniform distribution such that $R_{mq} \sim U(0,.3)$, ensuring that overall allele frequencies in each population were maintained. We then used the methods described above to estimate $P({S_{jq}})$ and $E(N_{S})$ and their variances given $\gamma = 30$. For comparison, we also conducted 10,000 random draws of size $\gamma$ from each loci in each population, then calculated $P({S_{jq}})$ and its variance for each locus empirically and $E(N_{S})$ by summing across all 100 loci for each set of draws. We likewise calculated the variance of $N_{S}$ directly across all sets of random draws.

To compare our calculated estimates to the empirical simulations, we used the implementation of the Agresti-Coull [@agrestiApproximateBetterExact1998] method from the R package `binom` [@dorai-rajBinomBinomialConfidence2022] to calculate 95% confidence intervals for each empirical $P({S_{jq}})$. We also used normal approximation to calculate 95% confidence and prediction intervals for $E(N_{S})$ and $N_{S}$ using $\sigma_{E(n_{S})}^{2}$ and the variance directly observed from the simulations, respectively.

@@

# Discussion
The number of segregating sites can provide an excellent counterpoint to measures of genetic diversity that are based on allele frequencies (such as observed and expected heterozygosity). For example, populations that have experienced a recent population expansion will often carry an excess of low frequency variants [@gattepailleInferringPopulationSize2013] caused by the recent increase in the overall rate at which mutations are produced in the population and the relative lack of time for any such new variants to drift to higher frequencies. The average expected heterozygosity across segregating sites may actually be lower (or at least lower than expected) in such cases than populations which have been demographically static, but they will carry far more segregating sites. A recent study in yellow perch provides an excellent example of such: several recently expanded populations show a relatively slight difference in heterozygosity in comparison to other, more demographically static populations but are segregating at far more loci [@schraidtDispersiveCurrentsExplain2023].
In cases like this, it is particularly important that the number of segregating sites is being properly calculated across populations, since failing to correct for differences in sample size or data missingness could mask biologically interesting signals of demographic history and obscure a critical facet of overall genomic diversity. The method we present here should therefore be a useful for future studies of genetic diversity.
